---
title: "Breast Cancer Wisconsin - Random Forest Classification Report"
author: "Data Science Report"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")

required_packages <- c("randomForest", "caret", "e1071", "knitr")
for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
}

library(randomForest)
library(caret)
library(knitr)

set.seed(42)
```

## 1. Data Ingestion

```{r data-ingestion}
bc_data <- read.csv("breast_cancer.csv", stringsAsFactors = FALSE)

bc_data$id <- NULL
empty_cols <- names(bc_data)[names(bc_data) == "" | grepl("^X$|^X\\.", names(bc_data))]
bc_data[empty_cols] <- NULL
bc_data <- bc_data[, colSums(is.na(bc_data)) < nrow(bc_data)]

numeric_cols <- sapply(bc_data, is.numeric)
for (col in names(bc_data)[numeric_cols]) {
  if (any(is.na(bc_data[[col]]))) {
    bc_data[[col]][is.na(bc_data[[col]])] <- median(bc_data[[col]], na.rm = TRUE)
  }
}

bc_data$diagnosis <- factor(bc_data$diagnosis, levels = c("B", "M"),
                             labels = c("Benign", "Malignant"))
bc_data <- bc_data[!is.na(bc_data$diagnosis), ]
```

```{r data-summary}
kable(data.frame(Metric = c("Rows", "Columns"),
                  Value = c(nrow(bc_data), ncol(bc_data))),
      caption = "Dataset Dimensions")

kable(as.data.frame(table(bc_data$diagnosis)),
      col.names = c("Diagnosis", "Count"),
      caption = "Class Distribution")
```

## 2. Data Partitioning

```{r partition}
set.seed(42)
train_index <- createDataPartition(bc_data$diagnosis, p = 0.80, list = FALSE)
train_data <- bc_data[train_index, ]
test_data  <- bc_data[-train_index, ]

kable(data.frame(Set = c("Training", "Test"),
                  Rows = c(nrow(train_data), nrow(test_data))),
      caption = "Train / Test Split (80/20)")
```

## 3. Model Fitting

```{r tune-mtry, results='hide'}
set.seed(42)
tuned_mtry <- tuneRF(
  x = train_data[, -which(names(train_data) == "diagnosis")],
  y = train_data$diagnosis,
  ntreeTry = 500,
  stepFactor = 1.5,
  improve = 0.01,
  trace = FALSE,
  plot = FALSE
)
best_mtry <- tuned_mtry[which.min(tuned_mtry[, "OOBError"]), "mtry"]
```

Selected `mtry`: **`r best_mtry`**

```{r train-model}
rf_model <- randomForest(
  diagnosis ~ .,
  data = train_data,
  ntree = 1000,
  mtry = best_mtry,
  importance = TRUE
)

saveRDS(rf_model, "rf_model.rds")

print(rf_model)
```

The trained model is saved to `rf_model.rds` in the working directory, and can be reloaded later without retraining:

```r
rf_model <- readRDS("rf_model.rds")
predict(rf_model, newdata = new_patient_data)
```

## 4. Model Evaluation

```{r evaluate}
rf_predictions <- predict(rf_model, newdata = test_data)

conf_matrix <- confusionMatrix(
  rf_predictions,
  test_data$diagnosis,
  positive = "Malignant"
)

saveRDS(conf_matrix, "confusion_matrix.rds")

conf_matrix
```

### Key Performance Metrics

```{r metrics-table}
metrics_df <- data.frame(
  Metric = c("Accuracy", "Sensitivity (Malignant Recall)", "Specificity"),
  Value = round(c(conf_matrix$overall["Accuracy"],
                   conf_matrix$byClass["Sensitivity"],
                   conf_matrix$byClass["Specificity"]), 4)
)
kable(metrics_df, caption = "Test Set Performance")
```

- **Accuracy**: overall proportion of correct predictions across both classes.
- **Sensitivity**: of all actual malignant cases, the proportion correctly flagged — the most clinically important number, since a missed malignant case is the costliest error.
- **Specificity**: of all actual benign cases, the proportion correctly cleared.

## 5. Feature Importance

```{r importance-table}
importance_df <- as.data.frame(round(importance(rf_model), 3))
kable(importance_df, caption = "Variable Importance Scores")
```

```{r importance-plot, fig.width=9, fig.height=7}
varImpPlot(
  rf_model,
  main = "Random Forest - Feature Importance",
  n.var = min(15, nrow(importance_df)),
  pch = 19,
  color = "steelblue"
)
```

## 6. Session Info

```{r session-info}
sessionInfo()
```
